BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 background.tex
Second Batch of fixes (Most of Daniele's)

Despite being extremely simple to implement, IAT patching suffers from a couple of disadvantages, which significantly limit its use in practice:
\begin{itemize}
    \item It is incredibly easy to detect by simply examining the entries of the IAT and checking whether or not each address falls inside the memory range of the DLL that should contain the function \cite{HookingDetection}.
    \item It is ineffective when function pointers are acquired dynamically, e.g. via \texttt{LoadLibrary} and \texttt{GetProcAddress} \cite{Buescher:2011:BIS:2186328.2186347}.
\end{itemize}

@@

EAT patching produces similar results to the ones obtained through IAT patching, but, unlike IAT patching, the created hooks are global, i.e. they affect every program which utilizes the altered DLL \cite{Berdajs:2010:EAU:1815744.1815746}.

@@

\begin{itemize}
\item Can be easily detected, for instance by comparing the code section of system libraries in memory with a matching original copy loaded from the file system to detect library modifications \cite{Buescher:2011:BIS:2186328.2186347} or by searching API entry points for specific patterns (e.g. presence of \texttt{jmp} instructions) \cite{HookingDetection}.
\item It needs additional modifications in the case where the function's entry points includes specific instructions, like ones which contain relative memory addresses. In fact, such instructions cannot be executed from a trampoline as the trampoline is located in a different memory location than the one of the original program code \cite{Berdajs:2010:EAU:1815744.1815746}.
\end{itemize}

\subsection{Debugger-Based Hooking}

@@

\section{Dynamic Binary Instrumentation}

@@

DBI, conversely from static binary instrumentation does not need any preprocessing of the program to be analyzed. This is what makes it more appealing for developers for profiling and tracing purposes. On top of that, DBI can be utilized for any program while static techniques are restricted in the code they can instrument, e.g. they cannot employed for code that is dynamically generated \cite{6658603}. The main drawback of DBI is that the instrumentation cost is encountered at run time, leading to performance degradation. However, in recent years, this issue has been mitigated by the introduction of generic DBI frameworks, which are accurately optimized to minimize the execution overhead \cite{Nethercote2004CI}.

>>>>

Despite being extremely simple to implement, IAT patching suffers from a couple of disadvantages, which significantly limit its use in practice:
\begin{itemize}
    \item It is incredibly easy to detect by simply examining the entries of the IAT and checking whether or not each address falls inside the memory range of the DLL that should contain the function \cite{HookingDetection}.
    \item It is ineffective when function pointers are acquired dynamically, e.g., via \texttt{LoadLibrary} and \texttt{GetProcAddress} \cite{Buescher:2011:BIS:2186328.2186347}.
\end{itemize}

@@

EAT patching produces similar results to the ones obtained through IAT patching, but, unlike IAT patching, the created hooks are global, i.e., they affect every program which utilizes the altered DLL \cite{Berdajs:2010:EAU:1815744.1815746}.

@@

\begin{itemize}
\item Can be easily detected, for instance by comparing the code section of system libraries in memory with a matching original copy loaded from the file system to detect library modifications \cite{Buescher:2011:BIS:2186328.2186347} or by searching API entry points for specific patterns (e.g., presence of \texttt{jmp} instructions) \cite{HookingDetection}.
\item It needs additional modifications in the case where the function's entry points includes specific instructions, like ones which contain relative memory addresses. In fact, such instructions cannot be executed from a trampoline as the trampoline is located in a different memory location than the one of the original program code \cite{Berdajs:2010:EAU:1815744.1815746}.
\end{itemize}

\subsection{Debugger-Based Hooking} \label{Debugger}

@@

\section{Dynamic Binary Instrumentation} \label{DBI}

@@

DBI, conversely from static binary instrumentation does not need any preprocessing of the program to be analyzed. This is what makes it more appealing for developers for profiling and tracing purposes. On top of that, DBI can be utilized for any program while static techniques are restricted in the code they can instrument, e.g., they cannot employed for code that is dynamically generated \cite{6658603}. The main drawback of DBI is that the instrumentation cost is encountered at run time, leading to performance degradation. However, in recent years, this issue has been mitigated by the introduction of generic DBI frameworks, which are accurately optimized to minimize the execution overhead \cite{Nethercote2004CI}.

### 2018-07-09 evaluation.tex
Second Batch of fixes (Most of Daniele's)

\iffalse
If JoeBox added differentiate machines
\fi
All the tests were conducted on a VirtualBox (version 5.2.6)  Virtual Machine with 1 CPU core and 3 GB of RAM running Windows 7 32-bit. The specifications of the host machine are:
\begin{itemize}
\item \textbf{Operating System:} Linux Mint 17.3 
\item \textbf{Processor:} Intel Core i7-3537U CPU @ 2.00 GHz $\times$ 2
\item \textbf{RAM:} 8 GBs  
\end{itemize}

@@

\begin{itemize}
\item \textbf{Empty Image.} All image notification functions are empty, that is, they just increase a global counter value. This means that no analysis routines are inserted at run-time. 
\item \textbf{Empty Routine.} All analysis routines are empty, i.e., all they do is, once again, increase a global counter value.
\item \textbf{Main Image.} Only Native APIs and APIs being directly invoked from the main executable of the pinned application are traced. This is BlueTracer's default mode of operation. 
\item \textbf{Complete.} Every event is traced, also including Native APIs and APIs being invoked outside the main executable.
\end{itemize}

@@

We tested each application three times and then recorded the average execution time. For the compression programs, the file being compressed was the Puppy Linux ISO (\texttt{xenialpup-7.5-uefi.iso}), of size 332 MBs. On the other hand, for the encryption and hashing utilities, we used the Ubuntu 18.04 ISO (\texttt{ubuntu-18.04-desktop-amd64}), the size of which is 1.8 GBs. During the tests, we also decided to stop the execution of the pinned application if, after it had been running for more than 10 minutes, the log file was greater than 5 GBs. This had to be done due to the hard drive limitations of the testing platform.

@@

\subsection{Anti-Debugging}

Anti-debugging checks are essentially aimed at determining if a program is running under a debugger. As already mentioned in section 2.1.5, there exist Windows APIs whose objective is exactly to discover if the calling process is being debugged.
A well-known API of this kind is \texttt{IsDebuggerPresent}, which returns a nonzero value if the calling process is running in the context of a debugger and zero otherwise. Such API is employed in one of Al-Khaser's anti-debugging checks and is appropriately recorded by BlueTracer (\textit{Listing 4.1}). As it can be seen below, the return value is zero, as no debugger is detected.

@@

Debugger detection is not only achieved through the use of Windows APIs. In fact, there are also some Native APIs which can be used for this purpose. One of them is \texttt{NtQueryInformationProcess}, which retrieves information related to the specified process. Its second parameter is utilized to specify the type of process information to be retrieved. When set to \texttt{ProcessDebugPort} (value \texttt{0x7}), the Native API provides information on whether or not the supplied process is being debugged. Specifically, if a debugger is present, the \texttt{ProcessInformation} output parameter is set to a non-zero port number, while, if the process is not under a debugger, this parameter is set to zero \cite{Sikorski:2012:PMA:2181153}. Al-Khaser performs this check and, once more, BlueTracer recorded it (\textit{Listing 4.3}). In particular, it can be deduced from \texttt{arg 1}'s value that \texttt{ProcessDebugPort} is adopted. Unfortunately, this time, we cannot tell from the log that a debugger has not been detected because the output paramater containing this information (\texttt{arg 2}) is a pointer to a \texttt{struct}, which changes based on the type of process information requested. BlueTracer does not yet address these complex cases and we are planning to address them in future versions of the tool.

@@

\subsection{Anti-Sandbox Timing-based}

@@

The \texttt{NtDelayExecution} Native API is lowest level user mode mechanism which can be employed to suspend execution. It takes as input two paramaters: a boolean that makes the Native API alertable when set to \texttt{true} and a \texttt{LARGE\_INTEGER} defining the delay interval. If the value for the second parameter is negative, it specifies the \textit{relative} amount to sleep in units of 100 nanoseconds; otherwise, it indicates an \textit{absolute time}, i.e. the date and time when to stop sleeping. Al-Khaser makes use of this Native API and BlueTracer was able to correctly trace it (\textit{Listing 4.4}). In particular, from the log entry, it can be seen that a negative delay was specified, meaning that a \textit{relative} time to sleep was provided.

@@

\subsection{Human Interaction Detection}

@@

\subsection{Anti-Virtualization}
Anti-virtualization checks look for a number of artifacts in the guest operating system introduced by hypervisors which are specific to virtual machines.

@@

\subsection{Anti-Analysis}

>>>>

\iffalse
If JoeBox added differentiate machines
\fi
All the tests were conducted on a VirtualBox (version 5.2.6)  Virtual Machine with 1 CPU core and 3 GB of RAM running Windows 7 32-bit. The specifications of the host machine are:
\begin{itemize}
\item \textbf{Operating System:} Linux Mint 17.3 
\item \textbf{Processor:} Intel Core i7-3537U CPU @ 2.00 GHz $\times$ 2
\item \textbf{RAM:} 8 GB  
\end{itemize}

@@

\begin{itemize}
\item \textbf{Empty Image.} All image notification functions are empty, that is, they just increase a global counter value. This means that no analysis routines are inserted at run-time. 
\item \textbf{Empty Routine.} All analysis routines are empty, i.e., all they do is, once again, increase a global counter value.
\item \textbf{Main Image.} Only Native APIs and APIs being directly invoked from the main executable of the pinned application are traced. This is BlueTracer's default mode of operation. 
\item \textbf{Complete.} Every event is traced, also including Native APIs and APIs being invoked outside the main executable (e.g., within libraries).
\end{itemize}

@@

We tested each application three times and then recorded the average execution time. For the compression programs, the file being compressed was the Puppy Linux ISO (\texttt{xenialpup-7.5-uefi.iso}), of size 332 MBs. On the other hand, for the encryption and hashing utilities, we used the Ubuntu 18.04 ISO (\texttt{ubuntu-18.04-desktop-amd64}), the size of which is 1.8 GB. During the tests, we also decided to stop the execution of the pinned application if, after it had been running for more than 10 minutes, the log file was greater than 5 GB. This had to be done due to the hard drive limitations of the testing platform.

@@

\subsubsection*{Anti-Debugging}

Anti-debugging checks are essentially aimed at determining if a program is running under a debugger. As already mentioned in Section~\ref{Debugger}, there exist Windows APIs whose objective is exactly to discover if the calling process is being debugged.
A well-known API of this kind is \texttt{IsDebuggerPresent}, which returns a nonzero value if the calling process is running in the context of a debugger and zero otherwise. Such API is employed in one of Al-Khaser's anti-debugging checks and is appropriately recorded by BlueTracer (\textit{Listing 4.1}). As it can be seen below, the return value is zero, as no debugger is detected.

@@

Debugger detection is not only achieved through the use of Windows APIs. In fact, there are also some Native APIs which can be used for this purpose. One of them is \texttt{NtQueryInformationProcess}, which retrieves information related to the specified process. Its second parameter is utilized to specify the type of process information to be retrieved. When set to \texttt{ProcessDebugPort} (value \texttt{0x7}), the Native API provides information on whether the supplied process is being debugged. Specifically, if a debugger is present, the \texttt{ProcessInformation} output parameter is set to a non-zero port number, while, if the process is not under a debugger, this parameter is set to zero \cite{Sikorski:2012:PMA:2181153}. Al-Khaser performs this check and, once more, BlueTracer recorded it (\textit{Listing 4.3}). In particular, it can be deduced from \texttt{arg 1}'s value that \texttt{ProcessDebugPort} is adopted. Unfortunately, this time, we cannot tell from the log that a debugger has not been detected because the output paramater containing this information (\texttt{arg 2}) is a pointer to a \texttt{struct}, which changes based on the type of process information requested. BlueTracer does not yet address these complex cases and we are planning to address them in future versions of the tool.

@@

\subsubsection*{Anti-Sandbox Timing-based}

@@

The \texttt{NtDelayExecution} Native API is lowest level user mode mechanism which can be employed to suspend execution. It takes as input two paramaters: a boolean that makes the Native API alertable when set to \texttt{true} and a \texttt{LARGE\_INTEGER} defining the delay interval. If the value for the second parameter is negative, it specifies the \textit{relative} amount to sleep in units of 100 nanoseconds; otherwise, it indicates an \textit{absolute time}, i.e., the date and time when to stop sleeping. Al-Khaser makes use of this Native API and BlueTracer was able to correctly trace it (\textit{Listing 4.4}). In particular, from the log entry, it can be seen that a negative delay was specified, meaning that a \textit{relative} time to sleep was provided.

@@

\subsubsection*{Human Interaction Detection}

@@

\subsubsection*{Anti-Virtualization}
Anti-virtualization checks look for a number of artifacts in the guest operating system introduced by hypervisors which are specific to virtual machines.

@@

\subsubsection{Anti-Analysis}
